Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-10-08 paper/manuscript.tex
Tweak the equilibration detection section
Also includes a tweak to references by @sboresch

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item non-equilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Karwounopoulos2024}.
\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine,Schoeller2023}
\end{itemize}

@@

A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Fig.~\ref{fig:automatic-equilibration-detection}).
Suppose we have a simulation of length $T$ consisting of correlated data.
Here, the goal of the post-processing step is to select the equilibration boundary $t_0 \in [0, T]$ so as to \emph{maximize} the number of effectively uncorrelated samples remaining in the production region $N_{[t_0,T]}$, which is defined as
\begin{eqnarray}
N_{[t_0,T]} &=& \frac{T - t_0}{g_{[t_0,T]}}
\end{eqnarray}
where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
Conveniently, this procedure also produces the information necessary to decorrelate the simulation data for estimating the free energy differences, a requisite next step in analysis.
This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.

@@

%MRS: Finlay, you can add information about RED here.

@@

If independent simulations at each $\lambda$ are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state. If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state. Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.

>>>>

While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues. 
Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item covalent inhibition~\cite{lameira2019predicting}
\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}. 
\item non-equilibrium free energy calculations~\cite{gapsys2020large}
\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Schoeller2023}.
\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine,Karwounopoulos2024}
\end{itemize}

@@

A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Fig.~\ref{fig:automatic-equilibration-detection}).
Suppose we have a simulation of length $T$ consisting of correlated data.
Here, the goal of the post-processing step is to select the equilibration boundary $t_0 \in [0, T]$ so as to \emph{maximize} the number of effectively uncorrelated samples remaining in the production region $N_{[t_0,T]}$, which is defined as
\begin{eqnarray}
N_{[t_0,T]} &=& \frac{T - t_0}{g_{[t_0,T]}}
\end{eqnarray}
where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice. A very similar approach involves minimising the standard error of the production region~\cite{whiteEffectiveTruncationHeuristic1997, sprattHeuristicsStartupProblem1998, hoadAutomatingWarmupLength2010, clarkRobustAutomatedTruncation2025}.

@@

@@

If independent simulations at each $\lambda$ are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state. If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state. Conveniently, maximising the effective sample size for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
